---
title: "Profit Analysis by Property"
author: "WRH Income Properties - Owned Properties"
date: "September 29, 2015"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
rm(list=ls()) # clear out objects
##setwd("c:/code/rfiles/rdata")
library("dplyr")
library("knitr")
library("ggplot2")
options(scipen=999)
wrh<- read.csv("currentdtl.csv")
md<-which(wrh$pno==280)#find Madalyn Landing - less than 12 months
wrh<- wrh[-md,]# remove Madalyn
row.names(wrh)<- wrh$pno
wrh$margin<- wrh$noiar/wrh$rev  # recalc margin just ot be sure
wrh$avgunitsize<- wrh$sqft/wrh$units
tot<- sapply(wrh[,c(2:19,21:26)],sum)#get totals for number columns
totmargin<-sum(wrh$noiar)/sum(wrh$rev)#get total margin
psqft<-cbind(wrh$pno,wrh[,2:19]/wrh$sqft,wrh$avgunitsize,wrh$margin) #create per square foot dataframe
colnames(psqft)[colnames(psqft)=="wrh$margin"]<-"margin"
colnames(psqft)[colnames(psqft)=="wrh$avgunitsize"]<-"AvgUnitSize"
colnames(psqft)[colnames(psqft)=="wrh$pno"]<-"prop#"
punit<-cbind(wrh$pno,wrh[,2:19]/wrh$units,wrh$avgunitsize,wrh$margin)#create per unit dataframe
colnames(punit)[colnames(punit)=="wrh$margin"]<-"margin"
colnames(punit)[colnames(punit)=="wrh$avgunitsize"]<-"AvgUnitSize"
colnames(punit)[colnames(punit)=="wrh$pno"]<-"prop#"
sumview<- c("gpr","rent","rev","totexp","noiar")
sview<- c("gpr","rent","rev","totexp","margin")
options(width=250)
```
This report analyzes the profitability of properties owned by WRH Income Properties. Profitability is defined as net operating income after replacements (NOIAR) divided by total revenue and it will be referred to in this report as margin. Debt service, corporate expenses and renovations are excluded in this analysis. Also, only properties with a 12 month operating history were included.  
I am presenting a number of views and summaries of the property results to shed some light on the margin levels and what affects them.  
To begin, here is list of all properties included in the analysis, sorted by their margin in decreasing order, followed by a histogram chart of the ranges of the margin.  

All information is based on trailing 12 month results for the period ended August 31, 2015.



```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE,fig.height=4,fig.width==8}
## comment =NA removes ## from output
#Margins by property
wrh$Margin<-format(wrh$margin,digits=3) #add a formatted margin - use a capital M
sum_margin<-wrh[,c("pno","prop","rev","noiar","Margin","margin")] #create a smaller dataframe
# t<- which (sum_margin$prop=="Total") # Find the Total Row to remove it
# sum_margin<-sum_margin[-t,] # Remove the Total Row
sum_margin<- sum_margin[order(sum_margin$Margin,decreasing = TRUE),]#Create a sorted DF
print(format(sum_margin[,2:5],justify="left",big.mark=","),print.gap=3) #Print it
print("Totals",quote="FALSE")
print(format(tot[c(6,18,23)],justify="left",big.mark=",",print.gap=3),quote="FALSE") #Print it
print ("Total Margin:",quote=FALSE)
print(totmargin,digits=3)

#boxplot(wrh$margin)
#hist(wrh$margin)
hist(wrh$margin, breaks=c(.2,.25,.30,.35,.40,.45,.50,.55,.6),main="Margins of WRH Properties",xlab="Margin(NOIAR/Revenue)",ylab="No. of Properties")
```

A summary by state is as follows:

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
 bystate<- group_by(wrh,state)
 bystate<- mutate(bystate,wtdmarg=rev/sum(rev)*margin)
 sumstate<-summarise(bystate, count=n(), revenue=sum(rev), noiar=sum(noiar), units=sum(units), sqfeet=sum(sqft))
 sumstate<- mutate(sumstate,marg=noiar/revenue)
 sumstate$marg<-format(sumstate$marg,digits=3)
 print(format(sumstate,justify="left",big.mark=","),print.gap=3)

```
The following is a summary of margin statistics and a segmenting of the properties into four groups of margin ranges (quartiles) from high to low:
```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
print ("Stat Summary of Margin%",quote=F)
print(summary(wrh$margin,digits=3))
```


```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
brk<-c(.25,.5,.75)
qnt<- quantile(wrh$margin,brk)
low<-wrh$margin<=qnt[1]
mid<-wrh$margin>qnt[1] & wrh$margin<=qnt[2]
high<-wrh$margin>qnt[2] & wrh$margin<=qnt[3]
topq<-wrh$margin>qnt[3]
#high<-wrh$margin>qnt[2] & wrh$margin<cutoff
#topq<-wrh$margin>=cutoff
print ("Top Quartile",quote=FALSE)
print(format(wrh[topq,c("prop","margin","units","sqft")],big.mark=",",digits=3))
print ("Third Quartile",quote=F)
print (format(wrh[high,c("prop","margin","units","sqft")],big.mark=",",digits=3))
print ("Second Quartile",quote=F)
print (format(wrh[mid,c("prop","margin","units","sqft")],big.mark=",",digits=3))
print ("Low Quartile",quote=F)
print (format(wrh[low,c("prop","margin","units","sqft")],big.mark=",",digits=3))
```

Following is a summary of average per square foot statistics and average per unit statistics.  
Average Per Square Foot:
```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}

labl<-c("low","mid","high","topq")
lowmean<-colMeans((psqft[low,sview]))
midmean<-colMeans((psqft[mid,sview]))
highmean<-colMeans((psqft[high,sview]))
topmean<-colMeans((psqft[topq,sview]))
meancat<-data.frame(low=lowmean)
meancat$mid<-midmean
meancat$high<-highmean
meancat$top<-topmean
print(meancat,digits=2)
```
Averarge Per Unit:

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}

labl<-c("low","mid","high","topq")
lowmeanu<-colMeans((punit[low,sview]))
midmeanu<-colMeans((punit[mid,sview]))
highmeanu<-colMeans((punit[high,sview]))
topmeanu<-colMeans((punit[topq,sview]))
meancatu<-data.frame(low=lowmeanu)
meancatu$mid<-midmeanu
meancatu$high<-highmeanu
meancatu$top<-topmeanu
print(format(meancatu,justify="left",big.mark=",",digits=2))

```

Here are the top five properties, followed by per unit and per square foot statistics:

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
#Identify Top 5 and create DF
cutoff<-min(sum_margin[1:5,"Margin"])
top5 <-which(wrh$margin>=cutoff)
wrhtop5<-wrh[top5,]
wrhtop5<-wrhtop5[order(wrhtop5$Margin,decreasing = TRUE),]
sum5<- wrhtop5[,c("prop","state","rev","noiar","units","sqft","Margin")]
sum5$revunit<-sum5[,3]/sum5[,5]
sum5$noiarunit<-sum5[,4]/sum5[,5]
sum5$revsqft<-sum5[,3]/sum5[,6]
sum5$noiarsqft<-sum5[,4]/sum5[,6]
sum5$Avg_unit_size<-sum5[,6]/sum5[,5]
print(format(sum5[,c(1,3,4,7,5)],justify="left", big.mark=","),print.gap=3) #Print it
```

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}

print(format(sum5[,c("revunit","noiarunit","revsqft","noiarsqft","Avg_unit_size")],justify="left", big.mark=","),print.gap=3) #Print it
```

To give a detail for basis of comaprison, here are all the properties' per unit statistics, sorted by margin from high to low

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}

  srt<-arrange(psqft[,c("prop#",sview)],desc(margin))
    print (format(srt,justify="left",big.mark=",", digits=3))
```

And some unit, property size and unit size statistics.
```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}

  propchar<-arrange(wrh[,c("prop","units","sqft","avgunitsize","mkt","margin")],desc(margin))
    print (format(propchar,justify="left",big.mark=",", digits=3))
```


Looking at this information, I reach the unsurprising conclusion that revenue per square foot has the most signifcant effect on the margin. It is not a pefect correlation, but it has a stronger relationship to margin than other measures.   Here is a plot of revenue per square foot to margin that shows the relationship between the two:

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
f<-lm(rev~wrh$margin, data = psqft)
with (psqft,plot(rev~wrh$margin,ylab="$rev/sqft",xlab="Margin"))
#lines(psqft$wrh$margin, fitted(f), col="blue")
#abline(h=10.6, col="red")
abline(f,col="blue")
```

Here are some take aways from this analysis:  
Six of the seven properties in the top margin quartile have revenue per square foot over $10. The only exception is Edenwood which has $8.37 of revenue per square foot but it has  very low expenses.  
The range of expenses per square foot is relatively small : $3.9 to $7.1 compared to the revenue range of $6.0 to $13.23 (see the following Box Plot).   This indicates most costs are fixed or semi fixed. While cost control is important,costs probably cannot be reduced enough to affect the margin substantially. The major lever to increase margin is revenue growth.

```{r, echo=FALSE,message=FALSE, warning=FALSE,comment=NA,row.names=FALSE}
    srt<-arrange(psqft[,c("prop#",sview)],desc(margin))
    #print (format(srt,justify="left",big.mark=",", digits=3))
    #summary(srt$rev)
    #summary(srt$totexp)
    boxplot(srt$totexp,srt$rev,names=c("OprExp$/Sqft","Rev$/SqFt"))

    
```


With that said, it is probably worthwhile to look at properties that are achieving revenue per square foot of $10 or more that have lower margins to determine if expense levels are in line with the property operations.   
